Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-08-15 text/sections/00-abstract.tex
adjusted writing

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one mechanism for worsening student outcomes at US public universities over this same time period.

>>>>

US public universities universities have systematically substituted away from tenure-track/tenured professors towards contingent lecturers for the last thirty years, while their states have systematically reduced the funding to these universities relative to enrolment.
This paper connects the changes the faculty composition at US public universities to the fall in the states' funding for their public universities, using a shift-share approach to instrument for state funding and identify the causal effects of stagnating funding for higher education.
A fall in funding of \$1,000 per student leads to univeristies employing 6 more lecturers on average; relative to enrolment, a fall of $10$\% in funding per student decreases the number of assistant professors per student by $1.4$\% and full professors by $1.2$\%, yet increases the number of lecturers per student by $4.4$\%.
Dynamic estimates show that these effects linger multiple years after the initial funding shock.
Analysis of all faculty at Illinois public universities shows that incumbent professors' salaries, promotion rate, and quit rate are all are not affected by funding cuts over 2011-2021, implying that the composition change arose by limiting the hiring of new tenure-track/tenured professors to replace leaving tenure-track/tenured professors.
These results show the institutional effects of stagnating state funding for higher education, and exhibit one possible mechanism for deteriorating student outcomes at US public universities over this same time period.

### 2023-08-15 text/sections/05-results.tex
adjusted writing

Incumbent professors are relatively unaffected by changes in the state funding for their university in the years following their hire.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 1\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero, so that faculty who are already at the university (incumbent faculty) seem to have no effects of state funding cuts passed on to them.

%These findings show that public universities increase (decrease) their count of tenure-track and tenured professors per student in years when revenues are more (less) plentiful, possibly by increasing hiring intensity or efforts to maintain incumbent professors.
%In the same vein, when funding increases, count of tenure-track and tenured professors per student increases --- and the opposite in years of funding cuts.
%These findings are in line with the observation that universities froze hiring for a couple of years in response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
%For lecturers, we see a negative effect which lines up with two trends noted in \autoref{sec:trends}: funding for public universities (per student) is decreasing while relative usage of lecturers increased substantially.

The composition of faculty changed at public universities 1990--2017 thanks to stagnating state funding, but there are little discernible effects on incumbent faculty.
This may imply that the rate that public universities hired faculty was impacted by state funding cuts.
I take recent data on the total faculty hired at US public universities over 2011--2021 (provided by \citealt{wapman2022quantifying}), to show that universities with lower state funding (per student) also hired fewer professors in the same time period --- shown in \autoref{fig:hiring-correlation}.
Similarly, the funding shock IV model shows that a $10$\% cut in per student state funding leads to $13$\% fewer faculty hires per student across the decade 2011--2021.\footnote{
    See \autoref{sec:appendix-hiring} for further details.
}
These results provide limited evidence that changes in faculty composition arose by disrupting hiring of new professors at public universities, corresponding to the results that incumbent faculty were not by meaningfully effected by the funding cuts.

>>>>

Incumbent professors are relatively unaffected by changes in the state funding for their university in the years following their hire.
\autoref{tab:faculty-shock-illinois-rolling} show estimates of a 1\% change in state funding per student on faculty salaries, rate of exiting the Illinois public university system, and rate of promotion.
The estimates are not discernible from zero, so that faculty who are already at the university (incumbent faculty) seem to have no effects of state funding cuts passed on to them.
Additionally, local projection estimates show that professors' salaries are also not affected on average in yeard following a funding shock  (\autoref{fig:salaries-illinois-lp-rolling}).
Lecturers on the other hand, see a 0.75\% increase in salary in the second and third year after a 10\% increase in state funding (\autoref{fig:salaries-lecturer-illinois-lp-rolling}), reflecting how lecturers are again more affected by changes in public university funding.
Similarly, promotion rate and exit rate among incumbent faculty  (from the Illinois public university system) are unaffected in the years after an initial funding shock (\autoref{fig:promoted-illinois-lp-rolling}, \ref{fig:exit-illinois-lp-rolling}).

These findings show that public universities increase (decrease) their count of tenure-track and tenured professors per student in years when revenues are more (less) plentiful.
In the same vein, when funding increased, count of tenure-track and tenured professors per student increased --- and the opposite in years of funding cuts.
These findings are in line with the observation that universities froze hiring for a couple of years in response to 2008 negative budget shocks, particularly for tenure-track positions \citep{turner2014impact}, and similarly from shocks to university endowments \citep{brown2014endowment}.
For lecturers, we see a negative effect which lines up with two trends noted in \autoref{sec:trends}: funding for public universities (per student) decreased while relative usage of lecturers increased substantially.

The composition of faculty changed at public universities 1990--2017 thanks to stagnating state funding, but there are little discernible effects on incumbent faculty.
This implied that changes in faculty compoisition arose by state funding cuts impacting the rate that public universioties hired new professors thanks.
I take recent data on the total faculty hired at US public universities over 2011--2021 (provided by \citealt{wapman2022quantifying}), to show that universities with lower state funding (per student) also hired fewer professors in the same time period --- shown in \autoref{fig:hiring-correlation}.
Similarly, the funding shock IV model shows that a $10$\% cut in per student state funding leads to $13$\% fewer faculty hires per student across the decade 2011--2021.\footnote{
    This limited sample of data are purel;y a cross-section for count of professor hires 2011--2021, so that local projection estimates are not possible here.
    See \autoref{sec:appendix-hiring} for further details.
}
These results provide limited evidence that changes in faculty composition arose by disrupting hiring of new professors at public universities, corresponding to the concurrent results that faculty composition changed while incumbent faculty were not by meaningfully affected by the funding cuts.

### 2023-08-15 text/sections/06-discussion.tex
re-writing
add LP estimates in IBHED data (no effects worth mentioning).
added LP estimates in IBHED data (no effects worth mentioning).
adjusted writing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Discussion section
\section{Discussion}

Universities received less state funding, relative to increasing enrolment, and reacted by increasing their reliance on lecturers relative to assistant and full professors in both the short and medium-run; these results imply that lecturers can be considered substitutes for professors in the tenure-system.\footnote{
    Additional results in Appendix \autoref{sec:appendix-substitution} formalises the marginal rate of substitution between lecturers and professors on tenure-track, showing how assistant and full professors were substituted for lecturers while there is no substitution between assistant and full professors.
}
A public university's state funding per student fell by an average of 36\%, and the number of lecturers per student increased by 113\% (fell by 17\% for assistant and 23\% full professors).
The estimates of elasticity for faculty count with respect to state funding imply that stagnating state funding explains about 40\% of the observed substitution towards lecturers and away from tenure-track and tenured faculty.

@@

While market forces have been effective in the successes of US higher education research, we should worry about the effects on education.
Enrolment in higher education has increased drastically since 1990, and public universities educate more than twice as many undergraduates in 2021 as their private counterparts.
%At the same time returns to higher education have increased drastically \citep{berman1998implications}.
Increasingly these students are being taught by lecturers, and not tenured faculty.
Lecturers are often employed on short-term or part-time contracts (adjunct), with limited job stability.
While we should worry for faculty, and their working conditions, there is credible evidence that relying on adjunct lecturers leads to worse student outcomes, relative to full-time lecturers and professors \citep{zhu2021limited}.
As tenure has increasingly become a private sector phenomenon, we should worry about the long-term effects on higher education, and the effect on the average student who attends a public university in the US.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Discussion section
\section{Discussion}
\label{sec:discussion}

%\textbf{implication: public unis are increasing sheer count of lecturers, in response to enrolment, and not their tenured faculty.
%SO prof per student is where the movemenet is most stark.}
Universities received less state funding, relative to increasing enrolment, and reacted by increasing their reliance on lecturers relative to assistant and full professors in both the short and medium-run; these results imply that lecturers can be considered substitutes for professors in the tenure-system.\footnote{
    Additional results in Appendix \autoref{sec:appendix-substitution} formalises the marginal rate of substitution between lecturers and professors on tenure-track, showing how assistant and full professors were substituted for lecturers while there is no substitution between assistant and full professors.
}
A public university's state funding per student fell by an average of 36\%, and the number of lecturers per student increased by 113\% (fell by 17\% for assistant and 23\% full professors).
The estimates of elasticity for faculty count with respect to state funding imply that stagnating state funding explains about 40\% of the observed substitution towards lecturers and away from tenure-track and tenured faculty.

@@

While market forces have been effective in the successes of US higher education research, we should worry about the effects on education.
Enrolment in higher education has increased drastically since 1990, and public universities educate more than twice as many undergraduates in 2021 as their private counterparts.
%At the same time returns to higher education have increased drastically \citep{berman1998implications}.
Increasingly these students are being taught by lecturers, and not tenured faculty.
Lecturers are often employed on short-term or part-time contracts (adjunct), with limited job stability.
While we should worry about the effect on faculty working conditions, there is credible evidence that relying on adjunct lecturers leads to worse education and student outcomes, relative to full-time lecturers and professors \citep{zhu2021limited}.
As tenure has increasingly become a private sector phenomenon, and the majority of US higher education students attend public institutions, we should worry about the long-term effects of stagnating funding for higher education in the US.

### 2023-08-15 text/sections/07-conclusion.tex
adjusted writing

This analysis investigates how the recent stagnation in state support for higher education has affected faculty, and their composition, at US public universities.
This work contributes to the literature along two primary dimensions.
Firstly, by isolating changes in state funding on public universities via state funding shocks, this work provides an explanation for the increased relaiance on lecturers and away from tenure-track and tenured professors at US public universities.
This approach used multiple methods to estimate the short- and medium-run effects of state funding cuts on faculty outcomes.
Secondly, this work asks investigates individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes in the Illinois university system.

I found that public universities have systematically substituted away from tenure-track and tenured professors, towards lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, while suggestive information shows falls in hiring, which together implies that the changes in faculty composition are driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

While costs education have been rising in the US, public universities have also dealt with declining state funding.
I show that these headwinds led to systematic change at public universities, changes that affect their faculty, likely limiting public universities in their goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards contingent lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty composition, research at public universities, and the impact on higher education as a whole.

>>>>

This paper investigates how the recent stagnation in state support for higher education has affected faculty, and their composition, at US public universities.
This work contributes to the literature along two primary dimensions.
Firstly, by isolating changes in state funding on public universities via state funding shocks, this work provides an explanation for the increased relaiance on lecturers and away from tenure-track and tenured professors at US public universities.
This approach used multiple methods to estimate the short- and medium-run effects of state funding cuts on faculty outcomes.
Secondly, this work asks investigates individual faculty are affected by changes in state funding for their university using a dataset new in the economic literature (IBHED).
These data allow for detailed analysis of thousands of faculty salaries, and employment outcomes in the Illinois university system.

Public universities have systematically substituted away from tenure-track and tenured professors, towards lecturers, in the face of persistent declines in state funding for higher education.
The effects on incumbent professors in the state of Illinois are non-distinguishable from zero, with suggestive evidence for falls in faculty hiring, implying that the changes in faculty composition are driven by reduced hiring of new professors.
Public universities are using more contingent lecturers to teach their students, while private universities continue to employ more tenure-track and tenured professors than their public counterparts, and each year the gap widens.

At the same time as higher education costs have been rising in the US, public universities have also dealt with declining state funding.
I show that these headwinds led to systematic change at public universities, changes that affect their faculty, and likely limiting public universities in their goals in research and education.
These results show large changes in faculty composition, and that stagnation in state support explains at least a third of the observed shift away from tenured professors and towards lecturers.
At the same time, private universities were not exposed to financial headwinds of the same magnitude or persistence.
While public universities continue to educate the majority of higher education students in the US, we should worry about the effects of restricting their funding has on faculty, and the wider impact on higher education as a whole.

### 2023-08-15 text/sections/10-appendix.tex
adjusted writing

% Start appendix
\section{Appendix}
\label{appendix}
This project used data which are fully public, and computational tools which are fully open-source.
As such, all code and data involved in this project are available at this project's Github repository, available at \url{https://github.com/shoganhennessy/state-faculty-composition}.
They may be used for replication, or as the basis for further work, as needed.
Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.

>>>>

% Start appendix
\section{Appendix}
\label{appendix}
This project used data which are fully public, and computational tools which are fully open-source.
As such, all code and data (anonymised versions where necessary) involved in this project are available at this project's Github repository, available at \url{https://github.com/shoganhennessy/state-faculty-composition}.
They may be used for replication, or as the basis for further work, as needed.
Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
